Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-28 paper_jasa.tex
defined visual p-value, tweaked equations and reference

This leads us to a definition for the $p$-value of a lineup: under the null hypothesis, each observer has a $1/m$ chance of picking the test statistic from the lineup.  For $K$ independent observers, let $X$ be  the number of observers picking the test statistic from the lineup. Under the null hypothesis $X \sim \text{Binom}({K, 1/m})$,  therefore:

\begin{dfn}\label{dfn:pvalue}
The $p$-value of a lineup of size $m$ is given as 
\[
P(X \ge x) = 1 - \text{Binom}({K, 1/m}) (x-1) = \sum_{i=x}^{K} {K \choose i} \left(\frac{1}{m}\right)^{i} \left(\frac{m-1}{m}\right)^{K-i},
\]
with $X$ defined as above. 
\end{dfn}
Note that for $x=0$ the $p$-value becomes --mathematically-- equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. If we increase either $m$ or $K$, we would be able to determine the value at a higher precision.

@@

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$, with an exact $x_{\alpha}$. For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$. Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. For Type II error, we need to calculate the chance that the observer sees that actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise, where $p_i$ is the probability that individual $i$ picks out the actual data plot from the lineup. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}(K, p)$. We can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot from the lineup.

@@

\red{Here is what my suggestion for notation \\  Use $B$ for logistic model and $\beta$ for linear model \\ Use $X$ to indicate logistic model covariate matrix and $U$ to denote binomial variable \\ $Y$ to indicate observed data. \\ Use $\pi_i$ instead of $p_i$ since $p_i$ is used for $p$-value.  \\ Should I change everything accordingly? I see that we have been careful not to use $\beta$ for power! }

@@

The model in equation \eqref{eqn:mixed} provides the subject and lineup specific probability of successful evaluation as 
\begin{equation} \label{eqn:mixed_power}
\hat p_{\ell i} = \frac{exp(\hat g_{\ell i})}{1+ exp(\hat g_{\ell i})}
\end{equation}
where $\hat g_{\ell i} = W_{\ell i} \hat {\delta} +  Z_{\ell i}  \hat {\tau}_{\ell i}$.

@@

Note that the \hh{left}  hand side in equation \eqref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$.

@@

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from equation \eqref{power_exp}. 
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is %almost as good as 
\hh{close to} the power of the conventional test.

@@

With multiple observers it is possible to estimate the strength of the signal, using the proportion of correct picks. From equation \eqref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
\[
\hat p_D= (1- {u}/{K})/({m-1})
\]
where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- observers find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that observers have a harder time detecting the actual data plot for even small $p$-values.

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$) estimated using Lemma \ref{lemma}. The solid line is a 45 degree line which represents the equity shown in equation \eqref{eqn:power_estimate}. p-values are shown on square root scale scale. \blue{Is it better now?}}
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

\blue{We also see this feature in figure \ref{fig:choice_reason}}.

>>>>

This leads us to a definition for the $p$-value of a lineup: under the null hypothesis, each observer has a $1/m$ chance of picking the test statistic from the lineup.  For $K$ independent observers, let $X$ be  the number of observers picking the test statistic from the lineup. Under the null hypothesis $X \sim \text{Binom}_{K, 1/m}$,  therefore:

\begin{dfn}\label{dfn:pvalue}
The $p$-value of a lineup of size $m$ is given as

\begin{equation}\label{binom}
P(X \ge x) = 1 - \text{Binom}_{{K, 1/m}} (x-1) = \sum_{i=x}^{K} {K \choose i} \left(\frac{1}{m}\right)^{i} \left(\frac{m-1}{m}\right)^{K-i}
\end{equation}

\noindent with $X$ defined as above, and $x$ is the number of observers selecting the actual data plot. 
\end{dfn}
Note that for $x=0$ the $p$-value becomes --mathematically-- equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. If we increase either $m$ or $K$, we would be able to determine the value at a higher precision.

@@

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$, with an exact $x_{\alpha}$. For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$. Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. For Type II error, we need to calculate the chance that the observer sees that actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise, where $p_i$ is the probability that individual $i$ picks out the actual data plot from the lineup. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$. We can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot from the lineup.

@@

\red{Here is what my suggestion for notation \\  Use $B$ for logistic model and $\beta$ for linear model \\ Use $X$ to indicate logistic model covariate matrix and $U$ to denote binomial variable \\ $Y$ to indicate actual data. \\ Use $\pi_i$ instead of $p_i$ since $p_i$ is used for $p$-value.  \\ Should I change everything accordingly? I see that we have been careful not to use $\beta$ for power! }

@@

The model in Equation \ref{eqn:mixed} provides the subject and lineup specific probability of successful evaluation as 
\begin{equation} \label{eqn:mixed_power}
\hat p_{\ell i} = \frac{exp(\hat g_{\ell i})}{1+ exp(\hat g_{\ell i})}
\end{equation}
where $\hat g_{\ell i} = W_{\ell i} \hat {\delta} +  Z_{\ell i}  \hat {\tau}_{\ell i}$.

@@

Note that the left hand side in Equation \ref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$.

@@

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. 
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is %almost as good as 
close to the power of the conventional test.

@@

With multiple observers it is possible to estimate the visual $p$-value, and a measure of plot signal strength. The visual $p$-value is calculated from the binomial formula (Equation \ref{binom}). Assuming $H_o$ is true, the probability of picking the actual data plot by chance is $1/m$, and the visual $p$-value is the probability that $x$ or more observers select that plot. Figure \ref{fig:pval_definition} displays this visual $p$-value for each lineup against the conventional $p$value.

Similary, the strength of the signal, is calculated using the proportion of observers picking actual data plot. It is closely related to the power of the test. From Equation \eqref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
\[
\hat p_D= (1- {u}/{K})/({m-1})
\]
where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- observers find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that observers have a harder time detecting the actual data plot for even small $p$-values.

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$). The solid line is a 45 degree line which represents the equity shown in Equation \eqref{eqn:power_estimate}. p-values are shown on square root scale scale. \blue{Is it better now?}}
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

\blue{We also see this feature in figure \ref{fig:choice_reason}}.
\hh{Mahbub, it's not directly fig \ref{fig:choice_reason} - it's actually the lineup corresponding to the large outlier at (0.2,  0.6) in the left panel of fig \ref{fig:choice_reason}.}
